---
title: "Hausarbeit ML-Seminar"
output: html_notebook
---

```{r setup}
# Load packages
library(tidyverse)
library(baseballr)
library(glmnet)
library(randomForest)
library(e1071) 
library(caTools) 
library(class) 

source("helper_functions.R")
source("ml_functions.R")
```

# Bang - Bang: Who can beat the Astros?
Here talk about Scandal


# Data
Talk about how it is being read in

and which pitcher we are using
```{r data}
# Getting data ----
# id_map <- data.table::fread("./data/player_id_map.csv") %>% 
#   rename("bbref_id" = BREFID)

# scherzer_data <- get_pbp_data("Max Scherzer", id_map)
# data.table::fwrite(scherzer_data, "./data/scherzer_data.csv")
scherzer_data <- data.table::fread("./data/scherzer_data.csv")

# hader_data <- get_pbp_data("Josh Hader", id_map)
# data.table::fwrite(hader_data, "./data/hader_data.csv")
hader_data <- data.table::fread("./data/hader_data.csv")

# cole_data <- get_pbp_data("Gerrit Cole", id_map)
# data.table::fwrite(cole_data, "./data/cole_data.csv")
cole_data <- data.table::fread("./data/cole_data.csv")

# darvish_data <- get_pbp_data("Yu Darvish", id_map)
# data.table::fwrite(darvish_data, "./data/darvish_data.csv")
darvish_data <- data.table::fread("./data/darvish_data.csv")

# verlander_data <- get_pbp_data("Justin Verlander", id_map)
# data.table::fwrite(verlander_data, "./data/verlander_data.csv")
verlander_data <- data.table::fread("./data/verlander_data.csv")

# kimbrel_data <- get_pbp_data("Craig Kimbrel", id_map)
# data.table::fwrite(kimbrel_data, "./data/kimbrel_data.csv")
kimbrel_data <- data.table::fread("./data/kimbrel_data.csv")
```

## How does the data cleaning work?
```{r cleaning}
clean_data <- scherzer_data %>% clean_pbp_data()
```

# Pitches
What are pitches, what can we see from pitch usage chart
```{r pitch-usage}
plot_pitch_usage(clean_data)
```

 Other pitch stats
```{r pitch-stats}
plot_pitch_movement(clean_data %>% filter(game_year > 2018))

plot_pitch_by_count(clean_data %>% filter(game_year > 2018))

plot_pitch_by_batter(clean_data %>% filter(game_year > 2018))

plot_pitch_velocity(clean_data %>% filter(game_year > 2018))
```
# Machine Learning
Describe concept and feature engineering done here
```{r ml-prep}
# Recoding Data ----
ml_data <- clean_data %>% filter(game_year > 2018) %>%  prep_for_ml()

outcome <- ml_data %>% select(pitch_type, is_fastball)

ml_ohe <- mltools::one_hot(
  data.table::as.data.table(ml_data %>% select(-pitch_type, -is_fastball))
  ) %>% 
  janitor::remove_constant() %>% 
  janitor::remove_empty()

ml_ohe$outcome <- factor(outcome$is_fastball)

split_data <- rsample::initial_validation_split(ml_ohe, c(0.9, 0.05))

training_data <- rsample::training(split_data) %>% 
  select(
    outcome,
    starts_with("freq"),
    starts_with("lag1"),
    starts_with("lag2")
    ) %>% 
  na.omit()

testing_data <- rsample::testing(split_data) %>% 
  select(
    outcome,
    starts_with("freq"),
    starts_with("lag1"),
    starts_with("lag2")
  ) %>% 
  na.omit()

validation_data <- rsample::validation(split_data) %>% 
  select(
    outcome,
    starts_with("freq"),
    starts_with("lag1"),
    starts_with("lag2")
  ) %>% 
  na.omit()
```

## Models
### Lasso
```{r lasso}
lasso_result <- run_lasso_regression(training_data, testing_data, validation_data)
```

### Ridge
```{r ridge}
ridge_result <- run_ridge_regression(training_data, testing_data, validation_data)
```

### Random Forest
```{r random-forest}
rf_result <- run_random_forest(training_data, testing_data, validation_data)
```

```{r rf-viz}
rf_imp <- get_predictor_importance_rf(rf_result$model, training_data)

plot(rf_imp$importance) +
  ggtitle("PFI") +
  xlab('Reduction in MSE (compared to average loss) when predictor included')

# Plotting
shap_values = shapviz::shapviz(rf_imp$shap)
shapviz::sv_importance(shap_values, show_numbers = T)
shapviz::sv_importance(shap_values, kind='bee')
treeshap::plot_contribution(rf_imp$shap, obs= 4)+
  ggtitle('SHAP Break-down: Sample no.5')+
  ylab('SHAP value')
```

### K-Nearest-Neighbor
```{r knn}
knn_result <- run_knn(training_data, testing_data, validation_data)
```

### Support Vector Machines
```{r svm}
svm_result <- run_svm(training_data, testing_data, validation_data)
```

## The winner is...


# Conclusion

